Title

### 2012-10-31 inc/este/este.tex
more of apl paper into este

%TODO consider moving this paragraph to chapter intro 
Sub-nanosecond (ultrafast) pulsed laser-driven electron sources are now being employed in a wide variety of scientific studies: in particular, ultrafast electron diffraction (UED) [1], ultrafast transmission electron microscopy (UEM) [2], and pulsed X-ray free electron lasers [3].
The spatial quality of these sources is often characterized in terms of their normalized transverse rms emittance [4,5]; $\varepsilon_{\smallT} = \Delta p_x \Delta x / (m_0 c)$, where $\Delta p_x = \sqrt{ \langle p_x^2 \rangle }$ is the rms transverse momentum of the source and $\Delta x = \sqrt{ \langle x^2 \rangle }$  is its size (cylindrical symmetry assumed), and $m_0$ is the electron rest mass and $c$ is the speed of light in vacuum.
As a reduction in the source size $\Delta x$ is limited either by the attainable diffraction-limited focusing or by the short-pulse Child's Law (screening of the electron gun's acceleration field) [6], significant improvements in the performance of scientific instruments employing pulsed electron sources through a reduction in their spatial emittance $\varepsilon_{\smallT}$ (and consequent increase in brightness [7]) may only be possible by decreasing the rms transverse momentum $\Delta p_x$ of the electron source – an intrinsic property of the emission source [4,5].

@@

>>>>

%TODO consider moving this paragraph to chapter intro 
Sub-nanosecond (ultrafast) pulsed laser-driven electron sources are now being employed in a wide variety of scientific studies: in particular, ultrafast electron diffraction (UED) [1], ultrafast transmission electron microscopy (UEM) [2], and pulsed X-ray free electron lasers [3].
The spatial quality of these sources is often characterized in terms of their normalized transverse rms emittance [4,5]; $\varepsilon_{\smallT} = \Delta p_x \Delta x / (m_0 c)$, where $\Delta p_x = \sqrt{ \langle p_x^2 \rangle }$ is the rms transverse momentum of the source and $\Delta x = \sqrt{ \langle x^2 \rangle }$  is its size (cylindrical symmetry assumed), and $m_0$ is the electron rest mass and $c$ is the speed of light in vacuum.
As a reduction in the source size $\Delta x$ is limited either by the attainable diffraction-limited focusing or by the short-pulse Child's Law (screening of the electron gun's acceleration field) [6], significant improvements in the performance of scientific instruments employing pulsed electron sources through a reduction in their spatial emittance $\varepsilon_{\smallT}$ (and consequent increase in brightness [7]) may only be possible by decreasing the rms transverse momentum $\Delta p_x$ of the electron source --- an intrinsic property of the emission source [4,5].

@@

To verify the efficacy of the experimental technique, we monitored the momentum distribution (Fourier plane spot size) for electrons emitted by 2PTE from gold using the $\sim$200fs green (523nm) $p$-polarized laser pulses.
Due to the 60$^{\circ}$ angle of incidence, the circular laser beam is focused by the 30cm focal length lens to an elliptical 50x100$\mu$m (half-width 1/e maximum (HWe$^{-1}$M) of the field) spot size on the photocathode surface.
The results, the measured Fourier plane beam size and electron yield, are displayed in Figure 2 as a function of the incident laser pulse energy, corrected to account for 10\% optical loss (mainly from the uncoated vacuum system windows).

The quadratic dependence of the electron yield on pulse energy clearly indicates a two-photon emission process.
This is expected since an excitation energy of two green photons ($\hbar \omega$ = 2.37eV) will be required to overcome the reported $\phi$ = 4.69eV effective work function of a thick gold film contaminated with adsorbed water [12] or ensure that at least the tail of the electron Fermi distribution has sufficient energy to overcome the 5.1($\pm$0.1)eV work function of a clean Au surface [13].
Moreover, when $2\hbar \omega \approx \phi$, the dependence of the yield on the square of the electron's excess energy above the work function [12] ensures that predominantly only the high energy (Boltzmann) tail of the Fermi distribution contributes to the emission.
Consequently, the observed increase in $\Delta p_x$ with incident pulse energy must be due to a heating effect; specifically, laser heating of the free electron Fermi gas as two-photon excitation is instantaneous and the $\sim$200fs green laser pulse duration excludes any coupling to the lattice which occurs on the time scale of a few picoseconds [14].
The solid line in Figure 2 is the predicted variation derived using a simple zero free parameter model of this effect.
Using the known optical properties of gold [8], which give a reflectance $R_p$ = 0.53 and an absorption depth of 20nm, and the temperature-dependent heat capacity of a free electron Fermi gas, the model evaluates the average temperature of the laser-heated electron gas that may be two-photon excited assuming that only electrons within a few nanometers of the surface can be emitted, as the mean free path for electrons in Au is $\sim$4nm [15].
This average electron temperature $T_e$ is then used in the AG model simulation of the experiment (Figure 1) to determine the expected Fourier plane spot size on the YAG scintillator; that is, using an initial $\Delta p_x = \sqrt{m_0 k_B T_e}$.  The dashed lines indicate the expected trends for a $\pm$10\% error in the laser spot size $\Delta x_0$ incident on the gold cathode surface.
The close agreement between the data and the simulation strongly supports our interpretation of the emission mechanism and, as expected, that the effective mass of a free electron in Au is equal to the rest mass $m_0$ [8].

Figure 3 depicts the results obtained for GaSb and InSb photocathode materials under pulsed UV (261nm) laser irradiation using the same experimental technique.
Both samples are cut from [100]-oriented polished wafers; the GaSb is undoped and the InSb is moderately $p$-type.
Prior studies [16] indicate that single-photon photoemission should not possible for either semiconductor --- the effective photoemission work functions (about 4.8eV for undoped GaSb and 4.89eV for $p$-type InSb (Fermi level pinned at valence band maximum)) being greater than the 4.75eV photon energy even when the 34meV Schottky barrier suppression due to the applied 8kV/cm DC field is included [4].
Nonetheless, for both zincblende semiconductors, significant laser-driven emission (more than expected for a single-photon field emission process) is observed with a yield nearly linearly proportional to the $\sim$4ps UV laser pulse energy.

The known optical properties of the two semiconductors [17] indicate a strong absorption at 261nm, with an optical absorption depth of only 7-8nm, which, based on the band structures of GaSb [18] and InSb [19], is primarily due to promotion of electrons near the $\Gamma$ point from the valence band (heavy-hole, light-hole, and split-off bands) directly into the upper $\Gamma_8$ conduction band located at 3.77eV and 3.59eV above the valence band maximum in GaSb and InSb, respectively.
Assuming parabolic bands and an estimated effective mass $m^*$ of about 0.3$m_0$ (0.5$m_0$) for the $\Gamma_8$ conduction band in GaSb (InSb), we determine that these electrons are excited with an average excess energy (above the $\Gamma_8$ conduction band minimum) of $\sim$0.35eV ($\sim$0.41eV).
For the $\sim$0.1nJ incident UV pulse energies, $\sim40\mu$m laser spot size, and $p$-polarized reflectance of $\sim$40\% for both semiconductors at our 60$^{\circ}$ incidence angle, the photo-injected electron density is $\sim10^{18}cm^{-3}$.
At these carrier densities, which are non-degenerate for the upper $\Gamma_8$ conduction bands, we expect rapid thermalization [20] and, consequently, an electron distribution with a well-populated Boltzmann tail extending above the vacuum level located at an effective work function $\phi$ of 0.99eV (1.18eV) above the $\Gamma_8$ conduction band minimum in GaSb (InSb); thus allowing for thermionic emission from this excited state as initially $\exp[\phi/(k_B T_e)] \approx 0.06$ in both semiconductors.

Carrier cooling, primarily by longitudinal optical (LO) phonon emission, and decay out of the upper conduction band to lower states will rapidly deplete the Boltzmann tail above $\phi$, thereby curtailing the observed ESTE.
By comparison with the $\Gamma_6$ conduction band Fr\"ohlich coupling constant in GaAs, where the characteristic LO phonon emission time is 165fs for a hot electron in the $\Gamma_6$ condunction band [21], we estimate that hot electrons in the upper $\Gamma_8$ conduction band of both GaSb and InSb emit a LO phonon every $\sim$200fs with an energy of 29 and 24meV, respectively.
This means that even without population decay mechanisms the temperature of the exited electron distribution $T_e$ drops at a rate of $\sim$1,600K/ps in both semiconductors.
We note that the close proximity of the lower and unconfined (in momentum space) $\Gamma_7$ upper conduction band to the $\Gamma_8$ band will likely result in a fast population decay, so that the observed ESTE process should have an intrinsic latency of much less than the $\sim$4ps UV laser pulse duration.

Figure 3 also shows that the measured Fourier plane electron beam spot size for GaSb is independent of the UV laser pulse energy, as would be expected for the proposed emission process.
More significantly, using $\Delta p_x = \sqrt{m_0 k_B T_e}$ (i.e., assuming the emitted electrons have a mass $m_0$ in the semiconductor), simulation of the experiment with the extended AG model [9] indicates that the observed GaSb Fourier plane spot size of 0.45($\pm$0.03)mm (HWe$^{-1}$M) would be due to an electron temperature $T_e \approx$ 360K which is associated with negligible thermionic emission ($\exp[-\phi/(k_B T_e)] \sim 10^{-15}$).
On the other hand, employing a mass of 0.3$m_0$ for the electrons in GaSb allows an average electron temperature (over the $\sim$4ps laser pulse duration) of about 1,200K to be extracted using the AG model simulation --- a value much more consistent with the proposed ESTE mechanism and the expected cooling rate by LO phonon emission.
In fact, at the maximum $\sim$0.2nJ incident UV laser pulse energy up to 108 electrons/pulse could be excited into the upper $\Gamma_8$ conduction band, of which one in $\exp[\phi/(k_B T_e)] \approx$ 20,000 are above the effective 0.99eV work function, thus agreeing with the observed yield of $\sim$10$^3$ electrons/pulse (Figure 3).
We also note that the normalized transverse rms emittance $\varepsilon_{\smallT}$ of this UV laser-driven ultrafast GaSb electron source is more than a factor of two less than that expected from a standard Cu photocathode ($\phi$ = 4.31eV and assuming $\Delta p_x = \sqrt{m_0 ( \hbar \omega - \phi ) / 3 }$, where $\hbar \omega$ is the incident photon energy [4,5]) irradiated at the same 261nm wavelength.
